---
title: "Slides and Plot"
author: "Andrew Hope"
date: "May 31, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Plotly Demo

```{r package, warning = FALSE, message = FALSE, echo = TRUE}
library(plotly)
library(dplyr)
```

## Plot Summary

This is a plot displaying the top 20 countries ranked by total alcohol consumption (in liters of pure alcohol per person per year). The bars are colored based on which type of alcohol (beer, spirits, wine) is consumed most within that country.

The data for this analysis comes from the GitHub repo of FiveThirtyEight.com:  
https://github.com/fivethirtyeight/data/tree/master/alcohol-consumption  

The original source for the data was The World Health Organisation, Global Information System on Alcohol and Health (GISAH), 2010.

## Slide with Plot

```{r data, cache = TRUE, warning = FALSE}
df <- read.csv("drinks.csv")
df <- df[order(df$total_litres_of_pure_alcohol, decreasing = TRUE),]
df.head <- df[1:20,]
df.head <- mutate(df.head, servings_lead = sapply(1:20, function(x) {order(c(df.head[[x, 2]], df.head[[x, 3]], df.head[[x, 4]]), decreasing = TRUE)[1]}))
df.head$servings_lead = cut(df.head$servings_lead, breaks = 3, labels = c("beer", "spirits", "wine"))
```

```{r plot, message = FALSE, warning = FALSE}
plot_ly(df.head, x = as.character(df.head$country), y = df.head$total_litres_of_pure_alcohol, mode = "markers", color = as.factor(df.head$servings_lead)) %>% layout(margin = list(l = 50, r = 50, b = 200, t = 50, pad = 4), yaxis = list(title = "Liters of Pure Alcohol (all types)"))
```