---
title: "Inference Analysis"
author: "Andrew Hope"
date: "April 6, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Inference Analysis Comparing Partitioned Data
## Overview
This analysis aims to determine if there is a significant difference in mean tooth length based on two different variables. The analysis will first describe the dataset, and will then use an independent two-sample t-test to look for significant differences between subset means. The anlysis will use an alpha value of 0.5, and it assumes that the samples were taken randomly from a population with an approximately normal distribution.

## Load Data
Load in the ToothGrowth dataset from the R dataset library.
```{r}
    library(datasets)
    tg <- ToothGrowth
```
## Exploratory Analysis and Transformations
Use some basic exploratory techniques to understand this small dataset.
```{r}
    dim(tg)
    summary(tg)
```
The quartiles for the dose column are a bit unusual in the summary. They are multiples of 0.5. Explore this further.
```{r}
    unique(tg$dose)
```
This variable should be a factor rather than numeric. Perform this transformation.
```{r}
    tg$dose <- as.factor(tg$dose)
    summary(tg)
```
## Calculate Statistics
Split the data by supp and dose.
```{r}
    supp <- split(tg$len, tg$supp)
    dose <- split(tg$len, tg$dose)
```
Calculate the mean, variance, and standard deviation for each value of supp. 
```{r}
    ms <- tapply(tg$len, tg$supp, mean)
    vs <- tapply(tg$len, tg$supp, var)
    ms
    vs
    sqrt(vs)
```
Do the same with dose.
```{r}
    md <- tapply(tg$len, tg$dose, mean)
    vd <- tapply(tg$len, tg$dose, var)
    md
    vd
    sqrt(vd)
```
## Compare Mean of Partitions
Compare the means (grouped by supp) using a 95% confidence interval and an independent two-tailed t-test.

```{r}
supp.t <- t.test(supp$OJ, supp$VC, conf.level = 0.95)
supp.t
```

For dose, execute an independent two-sample t-test between each factor level. This will result in three comparisons: 0.5 and 1, 0.5 and 2, and 1 and 2.  

First comparison -- factor level 0.5 to 1.
```{r}
dose.t12 <- t.test(dose$`0.5`, dose$`1`, conf.level = 0.95)
dose.t12
```

Second comparison -- factor level 0.5 to 2.
```{r}
dose.t13 <- t.test(dose$`0.5`, dose$`2`, conf.level = 0.95)
dose.t13
```

Third comparison -- factor level 1 to 2.
```{r}
dose.t23 <- t.test(dose$`1`, dose$`2`, conf.level = 0.95)
dose.t23
```

## Conclusions
For the supp variable, using a 95% confidence interval does not allow us to reject a null hypothesis. The two groups do not have significantly different means. This is indicated by the confidence interval containing zero, and also the p-value being greater than our accepted alpha of 0.5. 
```{r}
    supp.t$conf.int
    supp.t$p.value
```
For the dose variable, each comparison gave a confidence interval that did not include zero. This indicates that the groups have significantly different means. For all three comparisons, the p-value is less than 0.5, allowing us to comfortably reject the null hypothesis.
```{r}
    dose.t12$p.value
    dose.t13$p.value
    dose.t23$p.value
```