---
title: "Inference Analysis"
author: "Andrew Hope"
date: "April 6, 2018"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inference Analysis With an Exponential Distribution
## Overview
This analysis will demonstrate inferential statistics techniques to evaluate sampling mean and variance for a large sample with an underlying exponential distribution. The analysis will begin by creating a simulated dataset, and will then calculate and compare  sampling statistics. A final section provides plots to better convey the conclusions. 
```{r}
library(ggplot2)
```
## Simulations
Create a set of 1000 random samples from an exponential distribution with lambda equal to 0.2. Each sample has a size n = 40. Begin by generating all of the needed data points. In this analysis we round the number to 4 decimal places for the sake of keeping the data simpler and cleaner.
```{r generate}
r <- round(rexp(40000, rate = 0.2), 4)
```
Move the values into a matrix to create 1000 samples of size n = 40.
```{r matrix}
rmat <- matrix(r, 1000, 40)
head(rmat, 3)
```
## Sample Mean vs. Theoretical Mean
Find the mean for each of the 1000 rows, and take the mean of means to find the sampling mean.
```{r means}
sampling <- apply(rmat, 1, mean)
m <- mean(sampling)
m
```
The Central Limit Theorm allows us to conclude that this sampling mean is approximately equal to the theoretical mean. 

The theoretical mean of an exponential distribution is equal to 1/lambda. 
```{r}
1/0.2
```
Indeed, our sampling mean is very close to the theoretical mean.  

## Sample Variance vs. Thoretical Variance  
Calculate the population variance.  
```{r}
r.var <- var(r)
r.var
```
This should be close to the theoretical variance of an exponential distribution, which is equal to 1/lambda^2.
```{r}
1/0.2^2
```
Calculate the variance of the 1000 sample means to find the sampling variance.
```{r var}
v <- var(sampling)
v
```
This variance of the sample means should be approximately equal to its theoretical equivalent, (1/lambda^2)/n
```{r}
(1/0.2^2)/40
```
## Distribution
There is a significant difference between the distribution of an exponential random variable, and the distribution of means of a set of samples of exponential random variables.  

First, look at a distribution of a single exponential random variable.

```{r}
qplot(r, bins = 30)
```

Now, look at the distribution of the 1000 mean values from the simulated data set.

```{r}
qplot(sampling, bins = 30)
```

This demonstrates the Central Limit Threorem. The distribution of a sample's statistic, across a set of samples, will resemble a normal distribution.